iSEE: Interactive SummarizedExperiment Explorer

### 2018-03-29 manuscript/main.tex
Further nips and tucks, added relevant citations.

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

@@

\abstract{\textbf{Summary:} Data exploration is criticial to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualization are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor frameworks for analyzing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (https://bioconductor.org/packages/iSEE).\\
\textbf{Contact:} \href{aaron.lun@cruk.cam.ac.uk}{aaron.lun@cruk.cam.ac.uk}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\section{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics. 
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects, contaminants and low-quality samples. 
To this end, rapid visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, individual patients, single cells) and/or features (e.g., genes, transcripts, proteins, genomic regions).

Most existing tools for interactive visualization of biological data are designed for specific assays or analyses, e.g., pRoloc for proteomics data \citep{gatto2014mass}, shinyMethyl for methylation data \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
This makes it difficult to recycle the same visualization framework for new technologies or experimental designs where different aspects of the data are of interest. 
Moreover, standalone tools such as XXX do not easily integrate into established analysis frameworks such as those based on the R statistical programming language (CITE R Core Team). 
This complicates any coordinated use of these tools with a reproducible, transparent and rigorous statistical analysis.

\section{Description of the iSEE interface}
Here, we present the iSEE software package for interactive exploratory data analysis. 
iSEE is implemented in R using the interactive shiny framework and provides a number of features such as a customizable multi-panel layout, dynamic linking for point selection between plots, code reproducibility and intelligent downsampling. 
iSEE exploits the well-established SummarizedExperiment data architecture in the open-source Bioconductor project \citep{gentleman2004bioconductor}.
This enables iSEE to accommodate a wide range of `-omics datasets that are routinely processed by Bioconductor packages, ranging from conventional RNA sequencing to single-cell transcriptomics to mass cytometry.

\subsection{Integration with Bioconductor packages} 
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively. 
Per-feature or per-sample variables can also be easily added to the object in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results. 
The flexibility of the SummarizedExperiment class is the driving factor behind its wide deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C \citep{lun2016infrastructure}.
Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that stores sample-wise reduced dimensionality embeddings in single-cell `-omics data analyses. 
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization to complement existing analysis pipelines for a variety of data modalities. 
%In particular, we leverage off the state-of-the-art statistical and computational methods already available in the R/Bioconductor framework, allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

@@

\subsection{Customizable and flexibility}
iSEE also offers a host of other customizable features to improve data exploration. 
Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each individual  metadata variable and assay type consistently across all panels in the application. 
Row statistics tables can be augmented with dynamic annotation based on the selected row, providing links to online resources such as Ensembl or Entrez. 
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization. 
After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
iSEE also memorizes the exact R code that was used to generate every plot. 
This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.

\section{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets. 
We provide a number of case studies at (REPOSITORY NAME) to demonstrate how iSEE can be used to explore the TCGA RNA-seq data set (CITE someone?), the 68,000 PBMC single-cell RNA sequencing dataset from 10X Genomics (CITE Zheng) and a published mass cytometry dataset consisting of over 170,000 cells (CITE?). 
Interested users can also easily apply iSEE on their own data by downloading the package from Bioconductor (https://bioconductor.org/packages/iSEE).

>>>>

\documentclass{bioinfo}
\copyrightyear{2015} \pubyear{2015}
\access{Advance Access Publication Date: Day Month Year}
\appnotes{Manuscript Category}

@@

\abstract{\textbf{Summary:} Data exploration is criticial to the comprehension of large biological datasets generated by high-throughput assays such as sequencing.
However, most existing tools for interactive visualization are limited to specific assays or analyses.
Here, we present the iSEE (Interactive SummarizedExperiment Explorer) software package, which provides a general visual interface for exploring data in a SummarizedExperiment object.
iSEE is directly compatible with many existing R/Bioconductor packages for analyzing high-throughput biological data,
and provides useful features such as simultaneous examination of multiple data aspects, dynamic linking between plots and code tracking for reproducibility.
We demonstrate the utility and flexibility of iSEE by applying it to explore a range of real transcriptomics and proteomics datasets.\\
\textbf{Availability:} iSEE is publicly available as an R package from the open-source Bioconductor project (https://bioconductor.org/packages/iSEE).\\
\textbf{Contact:} \href{aaron.lun@cruk.cam.ac.uk}{aaron.lun@cruk.cam.ac.uk}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

@@

\section{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as transcriptomics. 
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples. 
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).

Most existing tools for interactive visualization of biological data are designed for specific assays and analyses, e.g., pRoloc for proteomics \citep{gatto2014mass}, shinyMethyl for methylation \citep{fortin2014shinymethyl}, HTSvis for high-throughput screens \citep{scheeder2017htsvis}.
Opportunities for customization are generally limited, making it difficult to re-use the same visualization software for new technologies or experimental designs where different aspects of the data are of interest. 
Moreover, standalone tools such as the Loupe Cell Browser from 10X Genomics \citep{zheng2017massively} do not easily integrate into established analysis pipelines such as those based on the R statistical programming language \citep{rcore2008R}. 
This complicates any coordinated use of these tools with a reproducible, transparent and statistically rigorous analysis.

Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits the widely-used SummarizedExperiment data architecture from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
This enables iSEE to accommodate a wide range of `-omics datasets that are routinely processed by Bioconductor packages, ranging from RNA sequencing to single-cell transcriptomics to mass cytometry.

\section{Description of the iSEE interface}

\subsection{Integration with Bioconductor packages} 
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively. 
Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results. 
The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that stores sample-wise reduced dimensionality embeddings for single-cell `-omics data analyses. 
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in the R/Bioconductor framework. 
%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

@@

\subsection{Customizing the user experience}
iSEE also offers a host of other customizable features to improve data exploration. 
Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each individual  metadata variable and assay type consistently across all panels in the application. 
Row statistics tables can be augmented with dynamic annotation based on the selected row, providing links to online resources such as Ensembl or Entrez. 
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization. 
After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
iSEE also memorizes the exact R code that was used to generate every plot. 
This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.

\section{Conclusion}
iSEE provides a general interactive interface for visual exploration of a diverse range of `-omics datasets. 
We provide a number of demonstrations at https://github.com/LTLA/iSEE2018, using iSEE to explore the TCGA RNA-seq collection (CITE?), the 68,000 PBMC single-cell RNA sequencing dataset from 10X Genomics \citep{zheng2017massively} and a published mass cytometry dataset consisting of over 170,000 cells (CITE?).
